Exploring mathematical objects from custom-tailored mathematical universes

### 2020-07-17 paper-filmat.tex
Final version

\documentclass[oneside,reqno]{amsart}

\usepackage[utf8]{inputenc}
\usepackage{amsthm,mathtools,amssymb,graphicx,ragged2e,extpfeil}
\usepackage{tikz}
\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing}
\usepackage{booktabs,framed}
\usepackage[all]{xy}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{xspace}
\usepackage{marginnote,animate,ocgx2}

\usepackage[natbib=true,style=numeric,maxnames=10]{biblatex}
\usepackage[babel]{csquotes}
\bibliography{paper-filmat.bib}

@@

\title[]{Exploring mathematical objects from custom-tailored mathematical universes}
\author{Ingo Blechschmidt}
\address{Universität Augsburg \\
Institut für Mathematik \\
Universitätsstr. 14 \\
86159 Augsburg, Germany}
\email{ingo.blechschmidt@math.uni-augsburg.de}

\theoremstyle{definition}
\newtheorem{defn}{Definition}[section]
\newtheorem{ex}[defn]{Example}

\theoremstyle{plain}
\newtheorem{prop}[defn]{Proposition}
\newtheorem{cor}[defn]{Corollary}
\newtheorem{lemma}[defn]{Lemma}
\newtheorem{thm}[defn]{Theorem}
\newtheorem{scholium}[defn]{Scholium}

\theoremstyle{remark}
\newtheorem{rem}[defn]{Remark}
\newtheorem{question}[defn]{Question}
\newtheorem{speculation}[defn]{Speculation}
\newtheorem{caveat}[defn]{Caveat}
\newtheorem{conjecture}[defn]{Conjecture}

@@

\newextarrow{\xbigtoto}{{20}{20}{20}{20}}{\bigRelbar\bigRelbar{\bigtwoarrowsleft\rightarrow\rightarrow}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\BB}{\mathbb{B}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\CC}{\mathbb{C}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\P}{\mathcal{P}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\op}{\mathrm{op}}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\Mod}{Mod}
\DeclareMathOperator{\Sh}{Sh}
\DeclareMathOperator{\Zar}{Zar}
\DeclareMathOperator{\PSh}{PSh}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\renewcommand{\_}{\mathpunct{.}\,}
\newcommand{\effective}{ef{}fective\xspace}
\newcommand{\effectively}{ef{}fectively\xspace}
\newcommand{\?}{\,{:}\,}
\newcommand{\realizes}{\Vdash}
\newcommand{\notnot}{\emph{not~not}\xspace}
\newcommand{\seq}[1]{\mathrel{\vdash\!\!\!_{#1}}}

@@

%\begin{abstract}
%  FUTURE: abstract
%\end{abstract}

\maketitle
\thispagestyle{empty}

@@

\begin{figure}[b]
  \tikzstyle{topos} = [draw=mypurple, very thick, rectangle, rounded corners, inner sep=5pt, inner ysep=10pt]
  \tikzstyle{title} = [fill=mypurple, text=white]

@@

We argue that toposes and specifically the change in perspective provided by
toposes are ripe for philosophical analysis. In particular, there are the
following connections with topics in the philosophy of mathematics:
\begin{enumerate}
\item Toposes enrich the realism/anti-realism debate in that they paint the larger
picture that the platonic heaven of mathematical objects is not unique: besides
the standard heaven of the standard topos, we can fathom the alternate
heavens of all other toposes, all embedded in a second-order heaven.
\item To some extent, the mathematical landscape depends on the commonly agreed-upon rules of
mathematics. These are not entirely absolute; for instance, it is conceivable
that from the foundational crisis Brouwer's intuitionism would have emerged as
the main school of thought and that we would now all reject the law of excluded
middle. Toposes allow us to explore alternatives to how history
has played out.
\item Mathematics is not only about studying mathematical objects, but also
about studying the relations between mathematical objects. The distinct view
on mathematical objects provided by any topos uncovers relations which
otherwise remain hidden.\footnote{Olivia Caramello's research
program~\cite{caramello:tst} provides a further topos-theoretic way for
uncovering hidden relations, though not between objects but between mathematical
theories. \textbf{\textcolor{red}{NOTE TO EDITOR: Insert cross reference to Olivia's chapter of
this book.}}}
\item In some cases, a mathematical relation can be expressed quite succinctly
using the language of a specific topos and not so succinctly using the language
of the standard topos. This phenomenon showcases the importance of
\emph{appropriate language}.
\item Toposes provide new impetus to study constructive mathematics and
intuitionistic logic, in particular also to restrict to intuitionistic
logic on the meta level and to consider the idea that the platonic heaven might
be governed by intuitionistic logic.
\end{enumerate}
We invite further research on these connections.

@@

Readers who would like to learn more details
are directed to the survey of category theory by Jean-Pierre
Marquis~\cite{sep:category-theory} and to a gentle introduction to topos theory
by Tom Leinster~\cite{leinster:introduction}. Standard references for the internal
language of toposes include~\cite[Chapter~VI]{moerdijk-maclane:sheaves-logic},
\cite[Chapter~14]{goldblatt:topoi},
\cite{caramello:preliminaries}, \cite{streicher:ctcl}, \cite{shulman:categorical-logic},
\cite[Chapter~6]{borceux:handbook3} and~\cite[Part~D]{johnstone:elephant}.

@@

In the seminal work introducing
toposes~\cite{artin-grothendieck-verdier:topos}, toposes are viewed as
generalized kinds of spaces. Every topological space~$X$ gives rise to a topos,
the \emph{topos of sheaves over~$X$}, and every continuous map gives rise to a
\emph{geometric morphism} between the induced sheaf toposes, but not every
topos is of this form. While the open sets of a topological space are required
to be parts of the space, the opens of toposes are not; and while for open
subsets~$U$ and~$V$ there is only a truth value as to whether~$U$ is contained
in~$V$, in a general topos there can be many distinct ways how an open is
contained in another one. This additional flexibility is required in situations
where honest open subsets are rare, such as when studying the étale cohomology
of a scheme~\cite{milne:etale}.

@@

A further perspective on toposes emerged in the early 1970s with the discovery
that toposes can be regarded as embodiments of a certain kind of first-order
theories, the \emph{geometric theories} briefly discussed on
page~\pageref{item:classifying-topos}. The so-called \emph{classifying toposes}
link geometrical and logical aspects and are fundamental to Olivia Caramello's
bridge-building program~\cite{caramello:tst}. Geometrically, the classifying
topos of a geometric theory~$\TT$ can be regarded as the generalized space of
models of~$\TT$; this idea is due to Monique
Hakim~\cite{hakim:relative-schemes}, though she did not cast her discovery in
this language. Logically, the classifying topos of~$\TT$ can be regarded as a
particular mathematical universe containing the \emph{generic~$\TT$-model}, a
model which has exactly those properties which are shared by all models.

Yet more views on toposes are fruitfully employed --
reference~\cite[pages~vii--viii]{johnstone:elephant} lists ten more -- but we shall not review
them here. A historical survey can be found in
reference~\cite{mclarty:history}.

\bigskip
\paragraph{Acknowledgments} We are grateful to Andrej Bauer, Martin
Brandenburg, Sina Hazratpour, Matthias Hutzler, Marc Nieper-Wißkirchen and
Alexander Oldenziel for invaluable discussions shaping this work, to Moritz
Laudahn, Matthias Hutzler and Milan Zerbin for their careful criticism of
earlier drafts and to Todd Lehman for the code for parts of
Figure~\ref{fig:landscape}. This note also profited substantially from the
thorough review by the three anonymous referees, whose efforts are much
appreciated and gladly acknowledged. We thank the organizers and all participants of the
Mussomeli conference of the Italian Network for the Philosophy in Mathematics,
where this work was presented, for creating an exceptionally beautiful meeting. In
particular, we are grateful to Neil Barton, Danielle Macbeth, Gianluigi Oliveri
and Lorenzo Rossi for valuable comments.

\tableofcontents

@@

\begin{defn}A \emph{topos} is a category which has all finite
limits, is cartesian closed, has a subobject classifier and contains a natural
numbers object.\footnote{More precisely, this is the definition of an
\emph{elementary topos with a natural numbers object}. Since this definition is
less tied to geometry than Grothendieck's (as categories of sheaves over
sites), there is a trend to call these toposes \emph{logoses}. However, that
term also has other uses.}\end{defn}

@@

\begin{defn}The \emph{standard topos}~$\Set$ is the category which has all
sets as its objects and all maps between sets as morphisms.\end{defn}

Given a topos~$\E$, we write~``$\E \models \varphi$'' to denote that a
mathematical statement~$\varphi$ \emph{holds in~$\E$}. The meaning of~``$\E \models
\varphi$'' is defined by recursion on the structure of~$\varphi$ following the
so-called \emph{Kripke--Joyal translation rules}. For instance, the rules for
translating conjunction and falsity read
\[ \begin{array}{lcl}
  \E \models (\alpha \wedge \beta) &\qquad\text{iff}\qquad&
  \E \models \alpha \quad\text{and}\quad \E \models \beta, \\[0.2em]
  \E \models \bot &\qquad\text{iff}\qquad&
  \text{$\E$ is the trivial topos}.
\end{array} \]
The remaining translation rules are more
involved~\cite[Section~VI.7]{moerdijk-maclane:sheaves-logic}; we do not list them here for
the case of a general topos~$\E$, but we will state them in the next sections
for several specific toposes. We refer to~``$\E \models \varphi$'' also as the
``external meaning of the internal statement~$\varphi$''.

@@

% FUTURE:
% \begin{rem}It might be helpful to picture mathematicians in a topos~$\E$. For
% exchanging ideas, they use the same mathematical language like we do, but their
% mathematical results seem unfamiliar or even wrong to us. Only when we remember
% to switch on the Kripke--Joyal translation engine their claims start making
% sense.\end{rem}

@@

\begin{thm}\label{thm:reasoning}Let~$\E$ be a topos and let~$\varphi$ be a
statement such that~$\E \models \varphi$. If~$\varphi$ intuitionistically
entails a further statement~$\psi$ (that is, if it is provable in
intuitionistic logic that~$\varphi$ entails~$\psi$), then~$\E \models
\psi$.\end{thm}

@@

\begin{defn}A topos~$\E$ is \emph{boolean} if and only if the laws of classical
logic are true in~$\E$.\end{defn}

@@

\begin{rem}The axiom of choice (which is strictly speaking not part of
classical logic, but of classical set theory) is also not available in most
toposes. By \emph{Diaconescu's theorem}, the axiom of choice implies the law of
excluded middle in presence of other axioms which are available in any topos.
\end{rem}

@@

In fact, to any model~$(M,\in)$ of a set theory such as~\textsc{zf}
or~\textsc{zfc}, there is a topos~$\Set_M$ such that a statement holds
in~$\Set_M$ if and only if it holds in~$M$.\footnote{The topos~$\Set_M$ can be
described as follows: Its objects are the elements of~$M$, that is the entities
which~$M$ believes to be sets, and its morphisms are those entities which~$M$
believes to be maps. The topos~$\Set_M$ validates the axioms of the structural
set theory \textsc{etcs}~\cite{mclarty:structuralism,marquis:foundations,barton-friedman:structures}, and models are isomorphic if and only if their
associated toposes are equivalent as
categories~\cite[Section~VI.10]{moerdijk-maclane:sheaves-logic}.}

\begin{ex}The topos~$\Set_V$ associated to the universe~$V$ of all sets (if
this structure is available in one's chosen ontology) coincides with the
standard topos~$\Set$.\end{ex}

@@

algebras)~\cite{butterfield-hamilton-isham:bohr,heunen-landsman-spitters:aqt}.

@@

theory)~\cite{lambek:incompatible,couture-lambek:reflections}.

@@

Toposes are still mathematical structures, and as long as we study
toposes within the usual setup of mathematics, our toposes are all part of the
standard topos. This is why Figure~\ref{fig:landscape} pictures the standard
topos twice, once as a particular topos next to others, and once as the universe
covering the entirety of our mathematical discourse.\footnote{There is a fine print to consider. Technically, if we
work within~\textsc{zf} or its intuitionistic cousin~\textsc{izf}, most toposes
of interest are proper classes, not sets. In particular~$\Set$ itself is a
proper class. Hence Figure~\ref{fig:landscape} should not be interpreted as
indicating that toposes are contained in~$\Set$ as objects, which most are not.
In this regard toposes are similar to class-sized inner models in set theory.
We believe that the vague statement ``our toposes are all part of the standard topos'' is still an
apt description of the situation. A possible formalization is (the trivial
observation that) ``our toposes are all indexed categories over~$\Set$''.}
The toposes which we can study in mathematics do not tell us
all possible stories how mathematics could unfold, only those which appear
coherent from the point of view of the standard topos, and the topos-theoretic
multiverse which we have access to is just a small part of an even larger
landscape.\footnote{This paragraph employs an overly narrow conception of
``mathematics'', focusing only on those mathematical worlds which form toposes
and for instance excluding any predicative flavors of mathematics
(Laura Crosilla's survey~\cite{crosilla:predicativity} is an excellent introduction).
Toposes are impredicative in the sense that any object of a topos is required to have a
powerobject. A predicative cousin of toposes are the \emph{arithmetic
universes} introduced by Joyal which have recently been an important object of
consideration by Maietti and
Vickers~\cite{maietti:au,maietti-vickers:induction,vickers:sketches}.}

@@

\begin{rem}The picture of a topos-theoretic multiverse is related to Hamkin's
multiverse view in set theory~\cite{hamkins:multiverse}. In fact, the
topos-theoretic multiverse can be regarded as an extension of the set-theoretic
multiverse: While Hamkins proposes to embrace all models of set theory (not
necessarily all of them equally -- we might prefer some models over others), we propose to embrace all toposes (again
not necessarily all of them equally). As every model~$M$ of set theory
gives rise to a topos~$\Set_M$, the set-theoretic multiverse is contained in the
topos-theoretic one.

However, a central and intriguing feature of the multiverse view in set theory
has, as of yet, no counterpart in topos theory: namely a systematic study of
its modal logic with respect to various notions of relations between toposes.\end{rem}

@@

Introductory literature on the \effective topos
includes~\cite{hyland:effective-topos,oosten:realizability,phoa:effective,bauer:c2c}.

\begin{defn}\begin{enumerate}
\item An \emph{assembly} is a set~$X$ together with a
relation~$({\realizes_X}) \subseteq \NN \times X$ such that for every element~$x
\in X$, there is a number~$n$ such that~$n \realizes_X x$.
\item A \emph{morphism of
assemblies}~$(X,{\realizes_X}) \to (Y,{\realizes_Y})$ is a map~$f : X \to Y$
which is \emph{tracked} by a Turing machine, that is for which there exists a
Turing machine~$M$ such that for any element~$x \in X$ and any number~$n$ such
that~$n \realizes x$, the computation~$M(n)$ terminates and~$M(n) \realizes
f(x)$.\end{enumerate}\end{defn}

@@

\begin{defn}The \emph{\effective topos}~$\Eff$ is the ex/reg completion (as
in~\cite[Section~3.4]{menni:exact}) of the category of assemblies.\end{defn}

@@

\begin{ex}\textbf{``Any natural number is prime or not.''} Even without knowing what
a prime number is, one can safely judge this statement to be true in
the standard topos, since it is just an instance of the law of excluded middle.

By the Kripke--Joyal semantics, stating that this statement is true in the \effective topos
amounts to stating that there is a Turing machine which, when given a natural
number~$n$ as input, terminates with a correct judgment whether~$n$ is prime or
not. Such a Turing machine indeed exists -- writing such a program is often a
first exercise in programming courses. Hence the statement is also true in the
\effective topos, but for the nontrivial reason that primality can be
algorithmically tested.
\end{ex}

\begin{ex}\textbf{``There are infinitely many primes.''} A first-order formalization
of this statement is ``for any natural number~$n$, there is a prime
number~$p$ which is greater than~$n$'', and is known to be true in the standard
topos by any of the many proofs of this fact.

Its external meaning when interpreted in the \effective topos is that there exists
a Turing machine~$M$ which, when given a natural number~$n$ as input, terminates with a
prime number~$p > n$ as output. Such a Turing machine exists, hence the
statement is true in the \effective topos.\footnote{More precisely, the
machine~$M$ should also output the description of a Turing machine which
witnesses that~$p$ is prime and that~$p > n$. However, the statement~``$p$ is prime and~$p > n$''
is~$\neg\neg$-stable (even decidable), and for those statements witnesses are
redundant.}
\end{ex}

\begin{ex}\textbf{``Any function~$\NN \to \NN$ is constantly zero or not.''} Precisely,
the statement is
\[ \forall f \? \NN^\NN\_
  \bigl((\forall n \? \NN\_ f(n) = 0) \vee
  \neg
  (\forall n \? \NN\_ f(n) = 0)\bigr). \]
By the law of excluded middle, this statement is trivially true in the standard
topos.

@@

A rigorous proof that such a machine~$M$ does not exist reduces its assumed
existence to the undecidability of the halting problem.
\end{ex}

>>>>

%\documentclass[oneside,reqno]{amsart}
\documentclass[graybox]{svmult}

\usepackage{type1cm}
\usepackage{graphicx}
\usepackage{multicol}
\usepackage[bottom]{footmisc}
\usepackage{newtxtext}
\usepackage{newtxmath}

\usepackage[utf8]{inputenc}
\usepackage{mathtools,ragged2e}
\usepackage{tikz}
\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing}
\usepackage{booktabs,framed}
\usepackage[all]{xy}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{xspace}
\usepackage{marginnote,animate,ocgx2}

\usepackage{natbib}

@@

%\theoremstyle{definition}
%\newtheorem{defn}{Definition}[section]
%\newtheorem{ex}[defn]{Example}
%
%\theoremstyle{plain}
%\newtheorem{prop}[defn]{Proposition}
%\newtheorem{cor}[defn]{Corollary}
%\newtheorem{lemma}[defn]{Lemma}
%\newtheorem{thm}[defn]{Theorem}
%\newtheorem{scholium}[defn]{Scholium}
%
%\theoremstyle{remark}
%\newtheorem{rem}[defn]{Remark}
%\newtheorem{question}[defn]{Question}
%\newtheorem{speculation}[defn]{Speculation}
%\newtheorem{caveat}[defn]{Caveat}
%\newtheorem{conjecture}[defn]{Conjecture}

@@

%\newextarrow{\xbigtoto}{{20}{20}{20}{20}}{\bigRelbar\bigRelbar{\bigtwoarrowsleft\rightarrow\rightarrow}}
\renewcommand{\E}{\mathcal{E}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\BB}{\mathbb{B}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\CC}{\mathbb{C}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\P}{\mathcal{P}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\op}{\mathrm{op}}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\Mod}{Mod}
\DeclareMathOperator{\Sh}{Sh}
\DeclareMathOperator{\Zar}{Zar}
\DeclareMathOperator{\PSh}{PSh}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\renewcommand{\_}{\mathpunct{.}\,}
\newcommand{\effective}{ef{}fective\xspace}
\newcommand{\effectively}{ef{}fectively\xspace}
\newcommand{\?}{\,{:}\,}
\newcommand{\realizes}{\Vdash}
\newcommand{\notnot}{\emph{not~not}\xspace}
\newcommand{\seq}[1]{\mathrel{\vdash\!\!\!_{#1}}}

@@

\title*{Exploring mathematical objects from custom-tailored mathematical universes}
\author{Ingo Blechschmidt}
\institute{Ingo Blechschmidt \at Universität Augsburg, Institut für Mathematik,
Universitätsstr. 14,
86159 Augsburg, Germany, \email{ingo.blechschmidt@math.uni-augsburg.de}}

\maketitle

\abstract{
  Toposes can be pictured as mathematical universes. Besides the standard topos,
  in which most of mathematics unfolds, there is a colorful host of alternate
  toposes in which mathematics plays out slightly
  differently.
  For instance, there are toposes in which the axiom of choice and the
  intermediate value theorem from undergraduate calculus fail.
  The purpose of this contribution is to give a glimpse of the toposophic
  landscape, presenting several specific toposes and exploring their peculiar
  properties, and to explicate how toposes provide distinct lenses through
  which the usual mathematical objects of the standard topos can be viewed.
  \emph{Key words:} topos theory, realism debate, well-adapted language,
  constructive mathematics
}
\medskip

@@

\begin{figure}[b]
  \centering
  \tikzstyle{topos} = [draw=mypurple, very thick, rectangle, rounded corners, inner sep=5pt, inner ysep=10pt]
  \tikzstyle{title} = [fill=mypurple, text=white]

@@

We argue that toposes and specifically the change in perspective provided by
toposes are ripe for philosophical analysis. In particular, there are the
following connections with topics in the philosophy of mathematics:
\begin{enumerate}
\item Toposes enrich the realism/anti-realism debate in that they paint the larger
picture that the platonic heaven of mathematical objects is not unique: besides
the standard heaven of the standard topos, we can fathom the alternate
heavens of all other toposes, all embedded in a second-order heaven.
\item To some extent, the mathematical landscape depends on the commonly agreed-upon rules of
mathematics. These are not entirely absolute; for instance, it is conceivable
that from the foundational crisis Brouwer's intuitionism would have emerged as
the main school of thought and that we would now all reject the law of excluded
middle. Toposes allow us to explore alternatives to how history
has played out.
\item Mathematics is not only about studying mathematical objects, but also
about studying the relations between mathematical objects. The distinct view
on mathematical objects provided by any topos uncovers relations which
otherwise remain hidden.\footnote{The research
program put forward by~\cite{caramello:tst} provides a further topos-theoretic way for
uncovering hidden relations, though not between objects but between mathematical
theories. \textbf{\textcolor{red}{NOTE TO EDITOR: Insert cross reference to Olivia's chapter of
this book.}}}
\item In some cases, a mathematical relation can be expressed quite succinctly
using the language of a specific topos and not so succinctly using the language
of the standard topos. This phenomenon showcases the importance of
\emph{appropriate language}.
\item Toposes provide new impetus to study constructive mathematics and
intuitionistic logic, in particular also to restrict to intuitionistic
logic on the meta level and to consider the idea that the platonic heaven might
be governed by intuitionistic logic.
\end{enumerate}
We invite further research on these connections.

@@

Readers who would like to learn more details
are directed to the survey of category theory by~\cite{sep:category-theory}
and to a gentle introduction to topos theory
by~\cite{leinster:introduction}. Standard references for the internal
language of toposes include~\cite[Chapter~VI]{moerdijk-maclane:sheaves-logic},
\cite[Chapter~14]{goldblatt:topoi},
\cite{caramello:preliminaries}, \cite{streicher:ctcl}, \cite{shulman:categorical-logic},
\cite[Chapter~6]{borceux:handbook3} and~\cite[Part~D]{johnstone:elephant}.

@@

In the seminal work introducing
toposes by~\cite{artin-grothendieck-verdier:topos}, toposes are viewed as
generalized kinds of spaces. Every topological space~$X$ gives rise to a topos,
the \emph{topos of sheaves over~$X$}, and every continuous map gives rise to a
\emph{geometric morphism} between the induced sheaf toposes, but not every
topos is of this form. While the open sets of a topological space are required
to be parts of the space, the opens of toposes are not; and while for open
subsets~$U$ and~$V$ there is only a truth value as to whether~$U$ is contained
in~$V$, in a general topos there can be many distinct ways how an open is
contained in another one. This additional flexibility is required in situations
where honest open subsets are rare, such as when studying the étale cohomology
of a scheme as in~\cite{milne:etale}.

@@

A further perspective on toposes emerged in the early 1970s with the discovery
that toposes can be regarded as embodiments of a certain kind of first-order
theories, the \emph{geometric theories} briefly discussed on
page~\pageref{item:classifying-topos}. The so-called \emph{classifying toposes}
link geometrical and logical aspects and are fundamental to Olivia Caramello's
bridge-building program set out in~\cite{caramello:tst}. Geometrically, the classifying
topos of a geometric theory~$\TT$ can be regarded as the generalized space of
models of~$\TT$; this idea is due to~\cite{hakim:relative-schemes}, though she did not cast her discovery in
this language. Logically, the classifying topos of~$\TT$ can be regarded as a
particular mathematical universe containing the \emph{generic~$\TT$-model}, a
model which has exactly those properties which are shared by all models.

Yet more views on toposes are fruitfully employed --
\cite[pages~vii--viii]{johnstone:elephant} lists ten more -- but we shall not review
them here. A historical survey was compiled by~\cite{mclarty:history}.

\bigskip
\paragraph{Acknowledgments} We are grateful to Andrej Bauer, Martin
Brandenburg, Sina Hazratpour, Matthias Hutzler, Marc Nieper-Wißkirchen and
Alexander Oldenziel for invaluable discussions shaping this work, to Moritz
Laudahn, Matthias Hutzler and Milan Zerbin for their careful criticism of
earlier drafts and to Todd Lehman for the code for parts of
Figure~\ref{fig:landscape}. This note also profited substantially from the
thorough review by three anonymous referees, whose efforts are much
appreciated and gladly acknowledged. We thank the organizers and all participants of the
Mussomeli conference of the Italian Network for the Philosophy in Mathematics,
where this work was presented, for creating an exceptionally beautiful meeting. In
particular, we are grateful to Neil Barton, Danielle Macbeth, Gianluigi Oliveri
and Lorenzo Rossi for valuable comments.

%\tableofcontents

@@

\begin{definition}A \emph{topos} is a category which has all finite
limits, is cartesian closed, has a subobject classifier and contains a natural
numbers object.\footnote{More precisely, this is the definition of an
\emph{elementary topos with a natural numbers object}. Since this definition is
less tied to geometry than Grothendieck's (as categories of sheaves over
sites), there is a trend to call these toposes \emph{logoses}. However, that
term also has other uses.}\end{definition}

@@

\begin{definition}The \emph{standard topos}~$\Set$ is the category which has all
sets as its objects and all maps between sets as morphisms.\end{definition}

Given a topos~$\E$, we write~``$\E \models \varphi$'' to denote that a
mathematical statement~$\varphi$ \emph{holds in~$\E$}. The meaning of~``$\E \models
\varphi$'' is defined by recursion on the structure of~$\varphi$ following the
so-called \emph{Kripke--Joyal translation rules}. For instance, the rules for
translating conjunction and falsity read
\[ \begin{array}{lcl}
  \E \models (\alpha \wedge \beta) &\qquad\text{iff}\qquad&
  \E \models \alpha \quad\text{and}\quad \E \models \beta, \\[0.2em]
  \E \models \bot &\qquad\text{iff}\qquad&
  \text{$\E$ is the trivial topos}.
\end{array} \]
The remaining translation rules are more
involved, as detailed by~\cite[Section~VI.7]{moerdijk-maclane:sheaves-logic}; we do not list them here for
the case of a general topos~$\E$, but we will state them in the next sections
for several specific toposes. We refer to~``$\E \models \varphi$'' also as the
``external meaning of the internal statement~$\varphi$''.

@@

% FUTURE:
% \begin{remark}It might be helpful to picture mathematicians in a topos~$\E$. For
% exchanging ideas, they use the same mathematical language like we do, but their
% mathematical results seem unfamiliar or even wrong to us. Only when we remember
% to switch on the Kripke--Joyal translation engine their claims start making
% sense.\end{remark}

@@

\begin{theorem}\label{thm:reasoning}Let~$\E$ be a topos and let~$\varphi$ be a
statement such that~$\E \models \varphi$. If~$\varphi$ intuitionistically
entails a further statement~$\psi$ (that is, if it is provable in
intuitionistic logic that~$\varphi$ entails~$\psi$), then~$\E \models
\psi$.\end{theorem}

@@

\begin{definition}A topos~$\E$ is \emph{boolean} if and only if the laws of classical
logic are true in~$\E$.\end{definition}

@@

\begin{remark}The axiom of choice (which is strictly speaking not part of
classical logic, but of classical set theory) is also not available in most
toposes. By \emph{Diaconescu's theorem}, the axiom of choice implies the law of
excluded middle in presence of other axioms which are available in any topos.
\end{remark}

@@

In fact, to any model~$(M,\in)$ of a set theory such as~\textsc{zf}
or~\textsc{zfc}, there is a topos~$\Set_M$ such that a statement holds
in~$\Set_M$ if and only if it holds in~$M$.\footnote{The topos~$\Set_M$ can be
described as follows: Its objects are the elements of~$M$, that is the entities
which~$M$ believes to be sets, and its morphisms are those entities which~$M$
believes to be maps. The topos~$\Set_M$ validates the axioms of the structural
set theory \textsc{etcs}, see~\cite{mclarty:structuralism,marquis:foundations,barton-friedman:structures}, and models are isomorphic if and only if their
associated toposes are equivalent as
categories, see~\cite[Section~VI.10]{moerdijk-maclane:sheaves-logic}.}

\begin{example}The topos~$\Set_V$ associated to the universe~$V$ of all sets (if
this structure is available in one's chosen ontology) coincides with the
standard topos~$\Set$.\end{example}

@@

algebras). Details are described by~\cite{butterfield-hamilton-isham:bohr,heunen-landsman-spitters:aqt}.

@@

theory). Details are described by~\cite{lambek:incompatible,couture-lambek:reflections}.

@@

Toposes are still mathematical structures, and as long as we study
toposes within the usual setup of mathematics, our toposes are all part of the
standard topos. This is why Figure~\ref{fig:landscape} pictures the standard
topos twice, once as a particular topos next to others, and once as the universe
covering the entirety of our mathematical discourse.\footnote{There is a fine print to consider. Technically, if we
work within~\textsc{zf} or its intuitionistic cousin~\textsc{izf}, most toposes
of interest are proper classes, not sets. In particular~$\Set$ itself is a
proper class. Hence Figure~\ref{fig:landscape} should not be interpreted as
indicating that toposes are contained in~$\Set$ as objects, which most are not.
In this regard toposes are similar to class-sized inner models in set theory.
We believe that the vague statement ``our toposes are all part of the standard topos'' is still an
apt description of the situation. A possible formalization is (the trivial
observation that) ``our toposes are all indexed categories over~$\Set$''.}
The toposes which we can study in mathematics do not tell us
all possible stories how mathematics could unfold, only those which appear
coherent from the point of view of the standard topos, and the topos-theoretic
multiverse which we have access to is just a small part of an even larger
landscape.\footnote{This paragraph employs an overly narrow conception of
``mathematics'', focusing only on those mathematical worlds which form toposes
and for instance excluding any predicative flavors of mathematics
(Laura Crosilla's survey in~\cite{crosilla:predicativity} is an excellent introduction).
Toposes are impredicative in the sense that any object of a topos is required to have a
powerobject. A predicative cousin of toposes are the \emph{arithmetic
universes} introduced by Joyal which have recently been an important object of
consideration by Maietti and
Vickers, see~\cite{maietti:au,maietti-vickers:induction,vickers:sketches}.}

@@

\begin{remark}The picture of a topos-theoretic multiverse is related to Hamkin's
multiverse view in set theory as put forward in~\cite{hamkins:multiverse}. In fact, the
topos-theoretic multiverse can be regarded as an extension of the set-theoretic
multiverse: While Hamkins proposes to embrace all models of set theory (not
necessarily all of them equally -- we might prefer some models over others), we propose to embrace all toposes (again
not necessarily all of them equally). As every model~$M$ of set theory
gives rise to a topos~$\Set_M$, the set-theoretic multiverse is contained in the
topos-theoretic one.

However, a central and intriguing feature of the multiverse view in set theory
has, as of yet, no counterpart in topos theory: namely a systematic study of
its modal logic with respect to various notions of relations between toposes.\end{remark}

@@

Introductory literature on the \effective topos
includes the references~\cite{hyland:effective-topos,oosten:realizability,phoa:effective,bauer:c2c}.

\begin{definition}\ \\[-1.2em]\begin{enumerate}
\item An \emph{assembly} is a set~$X$ together with a
relation~$({\realizes_X}) \subseteq \NN \times X$ such that for every element~$x
\in X$, there is a number~$n$ such that~$n \realizes_X x$.
\item A \emph{morphism of
assemblies}~$(X,{\realizes_X}) \to (Y,{\realizes_Y})$ is a map~$f : X \to Y$
which is \emph{tracked} by a Turing machine, that is for which there exists a
Turing machine~$M$ such that for any element~$x \in X$ and any number~$n$ such
that~$n \realizes x$, the computation~$M(n)$ terminates and~$M(n) \realizes
f(x)$.\end{enumerate}\end{definition}

@@

\begin{definition}The \emph{\effective topos}~$\Eff$ is the ex/reg completion (as
in~\cite[Section~3.4]{menni:exact}) of the category of assemblies.\end{definition}

@@

\begin{example}\textbf{``Any natural number is prime or not.''} Even without knowing what
a prime number is, one can safely judge this statement to be true in
the standard topos, since it is just an instance of the law of excluded middle.

By the Kripke--Joyal semantics, stating that this statement is true in the \effective topos
amounts to stating that there is a Turing machine which, when given a natural
number~$n$ as input, terminates with a correct judgment whether~$n$ is prime or
not. Such a Turing machine indeed exists -- writing such a program is often a
first exercise in programming courses. Hence the statement is also true in the
\effective topos, but for the nontrivial reason that primality can be
algorithmically tested.
\end{example}

\begin{example}\textbf{``There are infinitely many primes.''} A first-order formalization
of this statement is ``for any natural number~$n$, there is a prime
number~$p$ which is greater than~$n$'', and is known to be true in the standard
topos by any of the many proofs of this fact.

Its external meaning when interpreted in the \effective topos is that there exists
a Turing machine~$M$ which, when given a natural number~$n$ as input, terminates with a
prime number~$p > n$ as output. Such a Turing machine exists, hence the
statement is true in the \effective topos.\footnote{More precisely, the
machine~$M$ should also output the description of a Turing machine which
witnesses that~$p$ is prime and that~$p > n$. However, the statement~``$p$ is prime and~$p > n$''
is~$\neg\neg$-stable (even decidable), and for those statements witnesses are
redundant.}
\end{example}

\begin{example}\textbf{``Any function~$\NN \to \NN$ is constantly zero or not.''} Precisely,
the statement is
\[ \forall f \? \NN^\NN\_
  \bigl((\forall n \? \NN\_ f(n) = 0) \vee
  \neg
  (\forall n \? \NN\_ f(n) = 0)\bigr). \]
By the law of excluded middle, this statement is trivially true in the standard
topos.

@@

A rigorous proof that such a machine~$M$ does not exist reduces its assumed
existence to the undecidability of the halting problem.
\end{example}
